---
title: "Georgia Air Quality Trends: 2005-2015"
output: html_document
---



#A brief word about the software.

The software used in the data cleaning and analysis and the production of these graphics is R, a powerful open source statistacal software package that is rapidly becoming a standard for statistical analysis. It has the benefits of reproducibility as all analyses are scripted, so it is easy to replicate exactly the steps taken on a set of data, and even to quickly apply the same steps to new data. It is also highly flexible, with many addon packages available and maintained to aid in analysis and visualization. For more informatoni about R see <http://www.r-project.org/>. 

This report was generated using R Markdown and the Knitter Package which allows the code used in the analysis to be easily integrated directly into the report. Using the R Studio development environment and the markdown document. It is possible to run the code directly in the document.

The additional packages used in this research are 

```{r,message=FALSE}
require(plyr)
require(zoo)
require(ggplot2)
```

#Data Acquisition Initial Cleaning
```{r,cache=TRUE}
cat("Do you want to re-load the QData from the files? y/n \n
    Note! If the answer is n, the program will try loading it from a presaved RData file. If you are unsure, type y. \n
    Further Note! If you answer with y or an invalid response, the
    program will go ahead and load from files.")
runthis<-readLines(con=stdin(),n=1)
if(runthis!="n"){
   #Read in Data file names from "Data" folder
   files <- dir("Data")
   
   #Clear QData if it already exists to start with a blank slate
   if(exists("QData")){rm(QData)}
   
   #for every data file, read it in and add it to QData.
   for (file in files){
      
      #Read in the data files, pasting Data and the filename together for the path name
      QData_tmp <- read.csv(paste("Data",file,sep="/"),sep="|",header=TRUE,skip=0,na.strings="")
      
      #Data comes with two headers. Remove the Second Header
      QData_tmp <- QData_tmp[-1,]
      
      #Data comes with a comment at the end saying how many
      #rows were written, remove that too
      lastrow<-nrow(QData_tmp)
      QData_tmp <- QData_tmp[-lastrow,]
      
      #Check if QData exists. If it doesn't, use tmp data to make QData. If it does, add tmp data to QData.
      if (exists("QData")){
         QData <- rbind(QData,QData_tmp)
      }else{
         QData <- QData_tmp
      }
      print(paste(file,"loaded",sep=" "))
   }
   save(QData,file="QData.RData")
   
   rm(QData_tmp)
}

```

stuffystuff

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
